---
title: 'STAT 443: Lab 1'
author: 'Alex Grinius 20712972'
date: "11 January, 2025"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#if you do not have the package, type install.packages("name_of_the_package")
library(knitr)
library(tseries)
```

### Question 1

(a) 

```{r, echo=TRUE}
data <- read.csv("dat_Passengers.csv", header = TRUE)
head(data)
plot(data$DOMESTIC)
```

There is no clean way to show the date accocated with the data point as the month and year columns are split in 2


(b)

```{r, echo=TRUE}
time_series <- ts(data = data, start = 2002, frequency = 12)
```


(c)

```{r, echo=TRUE}
plot(time_series,plot.type = "single", xlab = "Years", main = "Air Passengers over time", ylab = "Passengers (TOTAL > DOMESTIC > INTERNATIONAL")
```
Here we can see the pandemic had a large effect on the number of passengers, much larger than the 2008 financial crisis. Number of passengers eems to be increasing over time, retuning to the trendline quickly after covid. We can also see periodicity in the number of passengers each year, having spikes over the summer and holiday season. 

### Question 2

(a) 

```{r, echo=TRUE}
set.seed(443)
X <- rnorm(500)
ts_x <- ts(data = X, start = 1, frequency = 1)

```


(b)

```{r, echo=TRUE}
plot(ts_x)
sum(abs(ts_x) > 1)
```
There are 157 data points that are more than one standard deviation away from the mean, which is close to the 33.3% we are looking for

(c)

```{r, echo=TRUE}
acf_x <- acf(ts_x)
```
This graph only has strong correlation at lag = 0, i.e. with itself. This is what we'd expect from a graph that was generated by purely random data. 